On Optimality of Finite Element Integration

### 2015-09-28 TOMS_Paper/assembly.tex
Rewrite the optimality proof

\usepackage{listings}
\lstset{language=C, breaklines=true, mathescape}

@@

Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former. In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}.

We start with introducing the transformation algorithm. Figure~\ref{code:intuition} provides the pseudo-code of the algorithm. An explanation and a discussion about its optimality follow.

\begin{figure}
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the transformation algorithm}
\label{code:intuition}
\end{figure}

Our choice is to study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''. If we estimate that, for a given monomial $M$, pre-evaluation will decrease the operation count, then: the sub-expression implementing $M$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.

@@

With the following Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach. First, however, note that because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space, with the search for the optimal by the algorithm being performed in a suitably constructed space.

\begin{Prop}
\label{prop:optimal-approach}
Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item the cost function $\theta$ is optimal. We define it such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, it returns two values $\theta_{w}$ and $\theta_{wo}$, representing the operation counts with and without pre-evaluation;
\item pre-evaluating different monomials does not result in identical tables;
\item there is no sharing of terms between monomials in $P$ and monomials in $Z$.
\end{enumerate} 
Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) We postpone the construction of the oracle $\theta$ to Section~\ref{sec:op_count}. 2) This is inherently a pathological case, which we neglect. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. We also neglect this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.

@@

A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$). We note that if all monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is trivially true. If, on the other hand, there is no (or only partial) sharing, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$. What we have to show now is that even by pre-evaluating more monomials, $\Lambda''_{ops_1} - \Lambda'_{ops_1} \geq I \cdot\delta$ holds. The intuition is that this should naturally be true since $\theta$ is optimal (assumption 1) and it predicted pre-evaluation to be a losing strategy for $Z_P$. However, to guarantee this we need to rely on assumption 2): if the new pre-evaluated tables were identical to those resulting from $P$, then we would have more sharing opportunities, which would generally break the optimality claim. Therefore, assumptions 1) and 2) are necessary for A) to be true.

B) Let us consider the non-trivial case: there are at least two monomials $p$ and $z$ that share terms, with $p \in P_Z$ and $z \in Z$. We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$ (both obvious due to assumption 1), we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle. The intuition behind this is that there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation. An extreme example is the summation of mass matrices, in which each monomial is characterized by same test and trial functions, but also by a different pre-multiplying coefficient (i.e., known) function. Assumption 3) enables us abstracting from such a scenario. 
\end{proof}

\subsection{Cost Function}
\label{sec:op_count}
It remains to tie one loose hand: the definition of the cost function $\theta = [\theta_{w}, \theta_{wo}]$. Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple expression tree traversal suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $\Lambda_{ops}^{ns}$. Assuming $I$ to be the size of $L_i$, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.

For $\theta_{w}$, things are more complicated. We first need to estimate the \textit{increase factor}, $\iota$, to account for the presence of (derivatives of) coefficients. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}.

@@

With a single coefficient, as we just saw, $\iota$ directly descends from the cost of product expansion. In general, however, predicting $\iota$ is less straightforward. For example, consider the case in which a monomial has multiple coefficients over the same function space. Product expansion would now lead to identical sub-expressions that, as such, would be mapped to the same temporary once pre-evaluated. The resulting loop nest would then be characterized by sharing, with the actual operation count (i.e., once sharing elimination is applied) being smaller than that expected by a naive analysis. For a precise evaluation of $\iota$, we need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).

>>>>

\usepackage{listings}
\lstset{language=C, breaklines=true, mathescape,
			 morekeywords={and, each}}

@@

Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former. In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}. For this, we start with explaining the transformation algorithm, which pseudo-code is provided in Figure~\ref{code:intuition}.

\begin{figure}
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial $m$:
  if $\theta(m)$ > 0 and memory constraints satisfied:
    mark $m$ as candidate for pre-evaluation
for each monomial $m$:
  if $m$ does not share terms with $m'$, an unmarked monomial:
    extract $m$ into a separate loop nest
    apply pre-evaluation to $m$
for each expression:
  remove sharing    
\end{lstlisting}
\caption{High level view of the transformation algorithm}
\label{code:intuition}
\end{figure}

Our choice is to study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''. We then introduce the cost function $\theta : M \rightarrow \mathbb{Z}$, where $M$ is the set of monomials in the form. $\theta(m)$ represents the change in operation count if pre-evaluation were applied to monomial $m$: a negative value indicates an increase, whereas a positive value indicates a gain. If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.

@@

Note that due to the memory constraints imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space in which the optimal is searched. We have defined them as high level operations on top of traditional expression-restructuring transformations such as code motion, re-association, and factorization. Other classes of loop transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.

Finally, with Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach.

\begin{Prop}
\label{prop:optimal-approach}
Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item pre-evaluating different monomials does not result in identical tables;
\item there is no sharing of terms between monomials in $P$ and monomials in $Z$.
\end{enumerate} 
Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
\end{Prop}
\begin{proof}
We first comment on the assumptions. First, we notice that neither of them is a sufficient condition for non-optimality. In addition: 1) is inherently a pathological case, which we neglect; 2) might actually occur in forms with several monomials, although, in practice, this was never the case in none of the problems that we have investigated (which are more than those studied in this article). Therefore, we also neglect this scenario, which would otherwise require a ``global analysis'' of $M$.

@@

A) By pre-evaluating the monomials in $Z_P$, we clearly have $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$, where $\delta > 0$ is a parameter to indicate the reduction in operation count exhibited by the non pre-evaluated monomials. To prove A), we then need to convince ourselves that $\Lambda''_{ops_1} -  I \cdot\delta \geq \Lambda'_{ops_1}$ holds. The only (corner) situation in which this might occur is whenever some of the pre-evaluated monomials in $Z_P$ result in tables identical to those already deriving from $P$. If that were the case, more sharing elimination opportunities would be exposed, which in turn could break the optimality claim. Assumption 1) prevents this situation, so the statement is true.

B) Let us consider the non-trivial case: there are at least two monomials $p$ and $z$ that share terms, with $p \in P_Z$ and $z \in Z$. We have to consider the fact that there could be (practically rare) circumstances in which despite $\theta_{wo}(z) < \theta_w(z)$ and $\theta_{w}(p) < \theta_{wo}(p)$, we could have $\theta^G_{wo}(z, p)  < \theta_w(p) + \theta_{wo}(z)$, where $\theta^G_{wo}$ is a hypothetical global oracle. The intuition behind this is that there could be a larger potential gain in sharing elimination that goes lost if monomials are inspected in isolation. An extreme example is a summation of mass matrices, in which each monomial is characterized by same test and trial functions, and by a different pre-multiplying coefficient (i.e., known) function. The previously discussed assumption 2) allows us to neglect this possibility, which then makes B) true. 
\end{proof}

\subsection{Cost Function}
\label{sec:op_count}
It remains to tie up one loose end: the definition of the cost function $\theta$. Since $\theta$ is expected to be used by a compiler to drive the transformation algorithm, requirements are simplicity and velocity.

We define $\theta(m) = \theta_{wo}(m) - \theta_{w}(m)$, where $\theta_{wo}$ and $\theta_{w}$ represent the operation counts for a monomial $m$ without and with pre-evaluation, respectively. In the following, we explain how to derive these two values.

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple expression tree traversal suffices to obtain the cost of a  sharing-free outer product loop nest, namely $\Lambda_{ops}^{ns}$. Assuming $I$ to be the size of $L_i$, we have that $\theta_{wo} = \Lambda_{ops}^{ns} \cdot I$.

Determining $\theta_{w}$ is more complicated. We first need to estimate the \textit{increase factor}, $\iota$, to account for the presence of (derivatives of) coefficients. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}.

@@

With a single coefficient, as we just saw, $\iota$ directly descends from the cost of product expansion. In general, however, predicting $\iota$ is less straightforward. For example, consider the case in which a monomial has multiple coefficients over the same function space. Product expansion would now lead to identical sub-expressions that, as such, would be mapped to the same temporary once pre-evaluated. The resulting loop nest would then be characterized by sharing, with the actual operation count (i.e., once sharing elimination is applied) being smaller than that found with a more naive analysis. For a precise evaluation of $\iota$, we need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).
